---
execute: 
  echo: true
  warning: false
---

# Data import, normalization, and exploratory analysis

In this chapter, you will learn how to read the transcript abundances
produced with _salmon_ [@patro2017salmon] to R, normalize data using
different approaches, and perform exploratory analysis on the data.
At the end of this chapter, we will be able to:

- read the output of _salmon_ [@patro2017salmon] to a `SummarizedExperiment`
object containing gene-level transcript abundances across samples.
- normalize data by library size and by cell (with spike-in standards)
- explore sample grouping with principal component analyses.

We will start by loading the required packages.

```{r here}
#| message: false
#| eval: true

library(here)
library(tidyverse)
library(rplaza)
library(fs)
library(tximeta)
library(SummarizedExperiment)
library(HybridExpress)
library(DESeq2)
library(patchwork)

set.seed(123) # for reproducibility
```

## Obtaining the example data

In this chapter, we will start from the output of __salmon__ [@patro2017salmon]
for all samples (2 genotypes, 3 ploidy levels; see Preface for details), 
produced with the _nfcore/rnaseq_ pipeline. _salmon_ returns one folder
per sample, each containing a `quant.sf` file with transcript abundances.

Let's download the example data from Zenodo and store it in a temporary
directory.

```{r}
# Download example data to a temporary directory
fpath <- file.path(tempdir(), "example_data.zip")
download.file(
    url = "https://zenodo.org/records/22938731/files/star_salmon.zip?download=1",
    destfile = fpath
)

# Unzip file
unzip(fpath, exdir = tempdir())
```

The output folder is called `star_salmon`. This is what it looks like:

```{r}
# Check the contents of the `star_salmon` folder
salmon_dir <- file.path(tempdir(), "star_salmon")

fs::dir_tree(salmon_dir)
```

We will also need a table of sample metadata, available in the same
Zenodo dataset.

```{r}
# Get sample metadata from Zenodo directly to R
coldata <- read_tsv(
    "https://zenodo.org/records/22938731/files/sample_metadata.tsv?download=1",
    show_col_types = FALSE
)

head(coldata)
```

Note that our metadata table includes variables describing ploidy level, 
genotype, genotype and ploidy (useful to extract model coefficients later on),
and presence/absence of spike-in standards (because spike-ins were not 
reliable in particular samples).

## Data import: from `quant.sf` files to a `SummarizedExperiment` object

To import data from `quant.sf` files, we will use the _tximeta_ package
[@love2020tximeta]. For that, we need to create a data frame with columns
`files` (paths to `quant.sf` files produced by salmon), and `names` (names
to be used as sample IDs).

```{r}
# Get paths to `quant.sf` files
qfiles <- list.files(
    salmon_dir, recursive = TRUE, pattern = "\\.sf", full.names = TRUE
)
head(qfiles)

# Create data frame required by {tximeta}
tximeta_df <-data.frame(files = qfiles, names = basename(dirname(qfiles)))
head(tximeta_df)

# Merge with `coldata` to include additional metadata variables
coldata <- left_join(tximeta_df, coldata, by = c("names" = "sample"))
head(coldata)
```

We can now import transcript abundances with the `tximeta()` function, which 
returns a `SummarizedExperiment` object with a count matrix of transcript 
abundances across samples, as well as sample metadata in the *coldata* slot.

```{r}
# Get `SummarizedExperiment object
se_tx <- tximeta(coldata, skipMeta = TRUE)
se_tx
```

Note that assays contain >54k features.
This is because transcriptomics quantifies **transcript abundances**, so our
estimates are at the transcript level. However, we're often interested in
abundances at the gene level (e.g., to find differentially expressed genes).
Hence, we will need to aggregate our transcript-level abundances to
gene-level abundances. To so do, we will need a correspondence table
between transcript IDs and gene IDs (i.e., transcripts and their parent genes).
There are many ways you can obtain such tables (e.g., from GTF files, 
from headers in the FASTA files for transcripts, etc). 
The PLAZA database [@van2022plaza] for comparative plant genomics
already provides such tables for all of its ~150 species, so we will retrieve
it from there.

```{r}
# Retrieve transcript-to-gene mappings from PLAZA
ath_tx2gene <- rplaza::get_id_conversions("ath") |>
    filter(alt_id_type == "tid") |>
    select(txid = alt_id, geneid = original_id)

head(ath_tx2gene)
```

If you have spike-in controls in your data, you will also need to add those
to the `tx2gene` table. In our data, feature IDs of spike-in standards
start with 'ERCC'.

```{r}
# Adding ERCC features (spike-ins) to `tx2gene`
tx2gene_ercc <- data.frame(txid = rownames(se_tx)) |>
    filter(str_detect(txid, "^ERCC")) |>
    mutate(geneid = str_replace_all(txid, "_gene", ""))

head(tx2gene_ercc)

tx2gene <- bind_rows(ath_tx2gene, tx2gene_ercc)

head(tx2gene)
tail(tx2gene)
```

We can then aggregate our transcript abundances to the gene level using
the `summarizeToGene()` function from _tximport_ [@soneson2016differential]. 
One of the main advantages of this method is that
it aggregates transcript-level abundances to gene-level abundances while
correcting for differences in transcript lengths (i.e., because longer
transcripts tend to have more reads mapped to them). However, since we have
3' tagged RNA-seq data, there are no differences in transcript lengths, because
we didn't sequence full transcripts. Hence, we will not correct for transcript
length biases here. We can import abundances with:

```{r}
# Aggregate abundances to the gene level
se <- summarizeToGene(
    se_tx, tx2gene = tx2gene, countsFromAbundance = "no", skipRanges = TRUE
)
se
```

We will also remove unnecessary assays and make sure our counts are integers
(required by _DESeq2_).

```{r}
# Remove `abundance` and `length` assays
assay(se, "abundance") <- NULL
assay(se, "length") <- NULL

# Round `counts` assay
assay(se, "counts") <- round(assay(se, "counts"))

se
```

This is what the count matrix and coldata (sample metadata) look like:

```{r}
assay(se, "counts") |> head()
colData(se) |> head()
```

If you're unfamiliar with `SummarizedExperiment` objects, please
have a look at the vignette of the `r BiocStyle::Biocpkg("SummarizedExperiment")`
package.

## Computing size factors: library sizes and cell volumes

Before estimating differential expression test statistics, the _DESeq2_ 
package [@love2014moderated] normalizes data using size factors
stored in the `sizeFactors` variable of the *colData* slot. If this column
doesn't exist in *colData*, _DESeq2_ will automatically add it using its
'median of ratios' method to estimate for robust library size estimation. Hence,
if we want to normalize our data using different approaches, all we need
to do is add a `sizeFactors` column to the *colData* slot of our 
`SummarizedExperiment` object, and _DESeq2_ takes care of the rest.
Here, we will demonstrate how to add size factors using three
approaches: (i) _DESeq2_'s median of ratios method, which corrects for 
differences in library sizes across samples; (ii) deconvolution, which is
a pooling-based approach that estimates size factors for each pool of samples 
(here, grouped by ploidy levels or ploidy+genotype), and then mathematically
deconvolves them back to sample-specific size factors; and (iii) using spike-in
standards to normalize data by cell volume.

We will first create alternative version of our data without spike-in
features.

```{r}
# Adding `sizeFactor` column to colData
## Method 1: Size factors with DESeq2's median of ratios
se_mr <- se[!grepl("ERCC", rownames(se)), ]
se_mr <- add_size_factors(se_mr, spikein = FALSE)

## Method 2: Size factors with spike-in controls
se_si <- se[, se$spike_in] # keep only samples with spike-ins
se_si <- add_size_factors(se_si, spikein = TRUE)

## Method 3: Size factors with deconvolution by `genotype_ploidy` level
se_de <- se[!grepl("ERCC", rownames(se)), ]
se_de$sizeFactor <- scuttle::pooledSizeFactors(
    se_de, clusters = se_de$genotype_ploidy
)
```

This is what size factors of each data set looks like:

```{r}
colData(se_mr)
colData(se_si)
colData(se_de)
```

Finally, we will filter out genes with too few counts across samples.
There's no universally good cut-off here, so we'd suggest playing with 
different values and seeing how many genes survive the filtering. Here,
we will remove genes with counts <1 in >=50% of the samples. To make the 
data analysis workflow simpler and avoid code repetition, we will combine 
these three objects into a list.

```{r}
# Create list of `SummarizedExperiment` objects
se_list <- list(MR = se_mr, SI = se_si, DE = se_de)

# Filter out genes with <1 count in >50% of the samples
se_list <- lapply(se_list, function(x) {
    n <- round(ncol(x) * 0.5)
    keep <- rowSums(assay(x) > 1) >=n
    x[keep, ]
})

se_list
```

With this cut-off, we go from ~32k genes to ~19k genes.

## Principal component analysis

We will now perform principal component analyses for each data set to
have a visual summary of our data and explore sample grouping. We will
first need to apply a variance-stabilizing transformation (to remove the
dependence between the mean and the variance). We will do so by using the 
`vst()` function from _DESeq2_.

```{r}
# Get variance-stabilized data
vsd_list <- lapply(se_list, function(x) {
    
    dds <- DESeqDataSet(x, design = ~ genotype + ploidy)
    vsd <- vst(dds, blind = TRUE)
    
    vsd
})
```

Don't worry too much about the `design` formula at this point. We've only 
included 'genotype + ploidy' because we want these two variables in the
PCA plot, but more details will come in the next chapter. We can now get
PCA plots for each data set. Note that PCA for transcriptomic data should never
be performed with the full data set. Instead, we use so-called
*highly variable features* to maximize biological signal. By default, the
`plotPCA()` from _DESeq2_ uses the top 500 genes with the highest variances,
but this can have different meanings depending on the genome size. Instead,
we'd suggest using a relative number (e.g., top 10% of the
genes with the highest variances), so that it at least scales with the
total number of genes.

```{r}
#| fig-width: 9
#| fig-height: 3

# Get PCA plot
p_pca <- lapply(vsd_list, function(x) {
    
    ## Get PCA data for top 10% genes with the highest variances
    pca_data <- plotPCA(
        x, intgroup = c("ploidy", "genotype"), returnData = TRUE, 
        ntop = round(nrow(x) * 0.1)
    )
    
    ## Plot using ggplot
    pct_var <- round(100 * attr(pca_data, "percentVar"))
    ggplot(pca_data, aes(x = PC1, y = PC2)) +
        geom_point(aes(color = ploidy, shape = genotype), size = 5) +
        theme_classic() +
        xlab(paste0("PC1: ", pct_var[1], "% variance")) +
        ylab(paste0("PC2: ", pct_var[2], "% variance")) +
        scale_color_manual(values = palette.colors()) +
        labs(color = "Ploidy", shape = "Genotype")
}) 

# Combine plots into a multi-panel figure
wrap_plots(
    p_pca$MR + labs(subtitle = "Median of ratios"),
    p_pca$DE + labs(subtitle = "Deconvolution"),
    p_pca$SI + labs(subtitle = "Spike-in"),
    nrow = 1
) +
    plot_layout(guides = "collect")
```

The figure shows that _DESeq2_'s median of ratios and the deconvolution-based
approach produce similar results, so we can simply choose one of them to
avoid replicating the same analysis 3 times. Expectedly, these
two methods produce different results compared to spike-in normalization,
so we will need to keep the spike-in-normalized data as a separate data set.

When normalizing by library size (with _DESeq2_ or deconvolution), we
see that PC1 mostly separates samples based on genotype. At the per cell level,
however, genotypes are separated in PC2, and PC1 doesn't seem to separate
samples by any of the variables we're interested in. In all cases, samples
from the same ploidy+genotype seem to group well together, although outliers
exist. It's also important to look at other PCs, as they might contain
additional information on sample grouping. To plot other PCs (e.g., PC1 vs PC3),
you'd need to modify the code above to include `pcsToUse = c(1, 3)` in 
`plotPCA()`.

## Wrapping up {.unnumbered}

We will now save `SummarizedExperiment` objects to `.rds` files for 
further reuse in the following chapters. Since _DESeq2_ normalization and 
deconvolution produce equivalent results, we will only save the former.

```{r}
#| eval: false

# Save `SummarizedExperiment` objects to RDS files
## DESeq2 normalization
saveRDS(
    se_list$MR, compress = "xz",
    file = here("products", "results", "se_deseq2.rds")
)

## Spike ins
saveRDS(
    se_list$SI, compress = "xz",
    file = here("products", "results", "se_spikein.rds")
)
```

## Session info {.unnumbered}

This document was created under the following conditions:

```{r sessioninfo}
#| eval: true
#| echo: false
sessioninfo::session_info()
```

## References {.unnumbered}
