---
execute: 
  echo: true
  warning: false
---

# Data import, normalization, and exploratory analysis

In this chapter, you will learn how to read the transcript abundances
produced with _salmon_ [@patro2017salmon] to R, normalize data using
different approaches, and perform exploratory analysis on the data.
At the end of this chapter, we will be able to:

- read the output of _salmon_ [@patro2017salmon] to a `SummarizedExperiment`
object containing gene-level transcript abundances across samples.
- normalize data by library size and by cell (with spike-in standards)
- explore sample grouping with principal component analyses.

We will start by loading the required packages.

```{r here}
#| message: false
#| eval: true

library(here)
library(tidyverse)
library(rplaza)
library(fs)
library(tximport)
library(SummarizedExperiment)

set.seed(123) # for reproducibility
```

## Obtaining the example data

In this chapter, we will start from the output of __salmon__ [@patro2017salmon]
for all samples (2 genotypes, 3 ploidy levels; see Preface for details), 
produced with the _nfcore/rnaseq_ pipeline. _salmon_ returns one folder
per sample, each containing a `quant.sf` file with transcript abundances.

Let's download the example data from Zenodo and store it in a temporary
directory.

```{r}
# Download example data to a temporary directory
fpath <- file.path(tempdir(), "example_data.zip")
download.file(
    url = "https://zenodo.org/records/22938731/files/star_salmon.zip?download=1",
    destfile = fpath
)

# Unzip file
unzip(fpath, exdir = tempdir())
```

The output folder is called `star_salmon`. This is what it looks like:

```{r}
# Check the contents of the `star_salmon` folder
salmon_dir <- file.path(tempdir(), "star_salmon")

fs::dir_tree(salmon_dir)
```

We will also need a table of sample metadata, available in the same
Zenodo dataset.

```{r}
# Get sample metadata from Zenodo directly to R
coldata <- read_tsv(
    "https://zenodo.org/records/22938731/files/sample_metadata.tsv?download=1",
    show_col_types = FALSE
) |>
    tibble::column_to_rownames("sample")

head(coldata)
```

Note that our metadata table includes variables describing ploidy level, 
genotype, genotype and ploidy (useful to extract model coefficients later on),
and presence/absence of spike-in standards (because spike-ins were not 
reliable in particular samples).

## Data import: from `quant.sf` files to a `SummarizedExperiment` object

To import data from `quant.sf` files, we will use the _tximport_ package
[@soneson2016differential]. For that, we need to specify the paths to
all `quant.sf` files, as well as a table of transcript-to-gene ID mappings.
This is because transcriptomics quantifies **transcript abundances**, but 
we're often interested in analyzing data at the gene level. Hence, we need
to aggregate transcript-level abundances to gene-level abundances.

There are many ways you can obtain correspondence tables between transcripts
and genes (e.g., from GTF files, from headers in the FASTA files for 
transcripts, etc). The PLAZA database [@van2022plaza] for comparative plant genomics
already provides such tables for all of its ~150 species, so we will retrieve
it from there.

```{r}
# Retrieve transcript-to-gene mappings from PLAZA
ath_tx2gene <- rplaza::get_id_conversions("ath") |>
    filter(alt_id_type == "tid") |>
    select(txid = alt_id, geneid = original_id)

head(ath_tx2gene)
```

Note that the order of the columns matter: transcript IDs must be in column 1,
and their corresponding parent gene IDs in column 2. We will need the paths
to all `quant.sf` files. This can be obtained with:

```{r}
# Define paths to `quant.sf` files
qfiles <- list.files(
    salmon_dir, recursive = TRUE, pattern = "\\.sf", full.names = TRUE
)

head(qfiles)
```

With that, we can import a table of gene-level transcript abundances 
across samples. One of the main advantages of using _tximport_ is that
it aggregates transcript-level abundances to gene-level abundances while
correcting for differences in transcript lengths (i.e., because longer
transcripts tend to have more reads mapped to them). However, since we have
3' tagged RNA-seq data, there are no differences in transcript lengths, because
we didn't sequence full transcripts. Hence, we will not correct for transcript
length biases here. We can import abundances with:

```{r}
# Import gene-level tx abundances across samples
counts <- tximport(
    qfiles, type = "salmon", tx2gene = ath_tx2gene, countsFromAbundance = "no"
)$counts

# Add proper column names (sample IDs)
colnames(counts) <- basename(dirname(qfiles))

# Make sure values are integers
counts <- round(counts)

# Taking a look at the matrix
head(counts)
```

In the Bioconductor ecosystem, package developers and users rely on 
the `SummarizedExperiment` class to store quantitative data and associated 
metadata. This ensures packages can seamless interoperate with each other,
and makes data analysis workflows simpler. We can create a 
`SummarizedExperiment` object with:

```{r}
# Making sure sample IDs in `coldata` match column names in `counts`
coldata <- coldata[match(rownames(coldata), colnames(counts)), ]

# Creating a `SummarizedExperiment` object
se <- SummarizedExperiment(
    assays = list(counts = counts),
    colData = coldata
)

se
```

In a `SummarizedExperiment` object, matrices are stored in the *assays* slot
(only one here, named `counts`), and column (i.e., sample) metadata is
stored in the *colData* slot. To access assays and colData, you can the 
following accessor functions:

```{r}
# Accessing first assay
assay(se, "counts") |> head()

#  Accessing colData 
colData(se) |> head()
```

## Computing size factors: library sizes and cell volumes





## Session info {.unnumbered}

This document was created under the following conditions:

```{r sessioninfo}
#| eval: true
#| echo: false
sessioninfo::session_info()
```

## References {.unnumbered}
